Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
committing changes to paper.qmd

#### State
The state ('state') variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Bar plot showing the distribution of observations across key battleground states in the 2024 United States Presidential Election polling data. The plot illustrates the count of polling observations for each state, highlighting states with more frequent polling activity. This visualization aids in understanding the level of engagement and data collection in each battleground state leading up to the election."

@@

#### End Date {#sec-end-date}
End date ('end_date') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling end dates for key battleground states in the lead-up to the 2024 United States Presidential Election. The histogram shows the frequency of polls conducted over time, providing insight into the timing of polling activities and potential trends in voter sentiment as Election Day approaches. This visualization is crucial for understanding how polling data correlates with the election timeline."

@@

#### Numeric Grade
Numeric Grade ('numeric_grade') represents the quality of each poll, as rated by [@FiveThirtyEight], based on factors such as methodology, transparency, and historical accuracy. In the context of our paper, numeric grade allows us to weigh the reliability of different polls. By including numeric grade as a variable in our model, we aim to account for variations in poll quality, ensuring that polls with higher ratings (indicating higher reliability) contribute more significantly to our predictions of voter support. @fig-numgrade-distribution displays the distribution of numeric grades for the polls included in our analysis, ranging from 2.5 to 3.0. Polls with higher grades are generally seen as more methodologically rigorous and transparent, which helps improve the reliability of our model’s predictions. The distribution shows several peaks, indicating that certain grade levels (around 2.7, 2.8, and 3.0) are more common among the polling data we collected.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Histogram illustrating the distribution of polling end dates for key battleground states in the lead-up to the 2024 United States Presidential Election. Displays the frequency of polls conducted over time, providing insight into the timing of polling activities and potential trends in voter sentiment as Election Day approaches. This visualization is crucial for understanding how polling data correlates with the election timeline."

@@

#### Transparency Score
Transparency Score ('transparency_score') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram illustrating the distribution of transparency scores for polls in key battleground states. Reveals the frequency of different transparency ratings, providing insight into the reliability of polling data. Understanding the distribution of transparency scores is crucial for evaluating the credibility of the polling results as Election Day approaches."

@@

#### Pollster
Pollster ('pollster') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar graph illustrating the distribution of polls by various pollsters in key battleground states. Displays the count of polls conducted by each pollster, highlighting the most frequently utilized polling organizations. Understanding the diversity of pollsters is essential for assessing the reliability and variance in polling data as Election Day approaches."

@@

#### Start Date
Start date ('start_date') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states in the lead-up to the 2024 United States Presidential Election. This visualization displays the frequency of polls conducted over time, providing insights into when polling activities were initiated and potential trends in voter sentiment as Election Day approaches. Understanding the timing of polls is crucial for evaluating the relevance of the polling data."

@@

The decision to utilize Bayesian logistic regression models stems from their flexibility and the ability to incorporate prior beliefs into the estimation process. This approach is particularly valuable in the context of electoral forecasting, where uncertainty is a significant factor. We defined the following priors for the model parameters:

- For the intercept $\beta_0$, we specified $\beta_0 \sim \text{Normal}(0, 10)$, providing a relatively uninformative prior that allows the data to inform the estimation of baseline log-odds.
- For the slope $\beta_1$, we used $\beta_1 \sim \text{Normal}(0, 5)$, reflecting reasonable expectations regarding the influence of polling trends over time.

Natural splines were chosen to model the effect of the end date due to the observed nonlinearities in voter support, allowing the model to adapt to changing public opinion. Key assumptions include the independence of observations within states and the appropriateness of the Beta distribution for modeling proportions. However, potential limitations include the reliance on the accuracy of polling data and the assumption that historical voting patterns are indicative of future behavior.

## Model Summary
### Trump Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

>>>>

#### State
The state ('state') variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground states in the lead-up to the 2024 United States Presidential Election. Notably, Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations. "

@@

#### End Date {#sec-end-date}
End date ('end_date') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states leading up to the 2024 United States Presidential Election, highlighting peaks in polling activity around late September."

@@

#### Numeric Grade
Numeric Grade ('numeric_grade') represents the quality of each poll, as rated by [@FiveThirtyEight], based on factors such as methodology, transparency, and historical accuracy. In the context of our paper, numeric grade allows us to weigh the reliability of different polls. By including numeric grade as a variable in our model, we aim to account for variations in poll quality, ensuring that polls with higher ratings (indicating higher reliability) contribute more significantly to our predictions of voter support. @fig-numgrade-distribution displays the distribution of numeric grades for the polls included in our analysis, ranging from 2.5 to 3.0. Polls with higher grades are generally seen as more methodologically rigorous and transparent, which helps improve the reliability of our model’s predictions. The distribution shows several peaks, indicating that certain grade levels (around 2.7, 2.8, and 3.0) are more common among the polling data we collected.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-numgrade-distribution
#| fig-cap: "Density plot illustrating the distribution of numeric grades assigned to pollsters within the polling data as the 2024 United States Presidential Election approaches. The graph highlights the frequency of numeric grades, indicating key concentrations that suggest variability in pollster quality. "

@@

#### Transparency Score
Transparency Score ('transparency_score') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data leading up to the 2024 United States Presidential Election. The plot indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

#### Pollster
Pollster ('pollster') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar chart displaying the distribution of polls conducted by various pollsters in key battleground states for the 2024 United States Presidential Election. The chart highlights the frequency of polling activities across different organizations, with Marist and Emerson standing out as the most prolific pollsters."

@@

#### Start Date
Start date ('start_date') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states leading up to the 2024 United States Presidential Election. The chart highlights the frequency of polling activities initiated over time, with a notable peak on September 14, indicating a surge in polling efforts as the election approached."

@@

The Bayesian Beta Regression model with a logit link function is the most suitable model for our prediction of the 2024 U.S. Presidential Election because it aligns closely with both the nature of the data and the goals of our paper, which include understanding voter support trends over time and accounting for regional differences across battleground states.

Our primary outcome variable is the percentage of support for each candidate in 7 key battleground states, scaled to a proportion between 0 and 1. Support percentages are naturally bounded between 0% and 100%, and this bounded nature poses challenges for traditional linear regression, which assumes an unbounded continuous outcome. The Beta distribution is ideal for modeling proportion data because it is inherently defined on the interval (0, 1). This allows us to model the proportions of support without the risk of generating unrealistic values outside this range. Unlike a linear model, which could predict support percentages exceeding 100% or falling below 0%, the Beta distribution respects the boundaries of our data. The logit link function further ensures that the model predicts in terms of log odds, which is suitable for interpreting support as probabilities—an essential consideration given our goal of forecasting potential victory in each state.

One of our key goals is to examine how support for each candidate evolves over time as Election Day approaches. Polling data often show non-linear trends, with fluctuations in support that may correspond to campaign events, debates, or shifts in the political landscape. Using natural splines with end_date_num (the number of days since the earliest poll date) allows the model to capture these non-linear trends in a flexible way. By including natural splines with 4 degrees of freedom, we allow the model to capture gradual changes in voter sentiment while avoiding overfitting to noise or abrupt fluctuations in the data. This is particularly important in the context of election polling, where voter opinions can change either slowly or rapidly depending on unpredictable external events. Alternative models, such as standard logistic regression without splines, would fail to capture these changes, potentially leading to a less accurate depiction of how support varies over time.

A unique aspect of our analysis is the focus on battleground states, which have distinct political dynamics that can heavily influence election outcomes. Incorporating state as a random effect provides flexibility in estimating state-specific baseline support levels for each candidate, while still allowing us to identify overall trends across all states. This approach is more effective than treating state as a fixed effect because it avoids overparameterizing the model with individual state coefficients, which could lead to overfitting in cases with limited data for some states. The random effect structure also helps generalize the model’s predictions, making it more robust when predicting support in battleground states with varying polling frequencies.

The Bayesian approach offers advantages for our model by allowing us to incorporate prior information, enhancing the stability and reliability of our estimates. Given that polling data can be sparse for certain states or dates, Bayesian inference helps stabilize parameter estimates, especially when there are limited data points for certain states or specific dates. By using weakly informative priors, we let the data primarily determine the model’s outcomes, while still providing some structure to prevent overfitting or extreme parameter values. The Bayesian framework also produces credible intervals, which provide a clear interpretation of uncertainty around our predictions. In the context of election forecasting, this is valuable as it enables us to quantify the range of plausible outcomes for each candidate’s support, adding a layer of transparency to our predictions.
This model provides interpretability that aligns well with the goals of our paper. By structuring the model with a logit link function, we can interpret the results in terms of odds or probability of victory, which is highly relevant in an election context. This interpretability allows us to understand the likelihood of each candidate winning in a specific state at a given time.

Additionally, the Bayesian Beta regression model offers flexibility in how we examine each candidate’s trends individually. By constructing separate models for Trump and Harris, we can directly compare their temporal trends and state-level variations without confounding effects.

Overall, the Bayesian Beta regression model with a logit link function is the most suitable model for our study. It allows for flexible non-linear trends over time, accounts for regional variation across battleground states, and provides interpretable results in terms of victory probabilities. The Bayesian framework stabilizes estimates in the presence of sparse data and provides meaningful uncertainty estimates. The model’s structure aligns well with our goal of forecasting each candidate’s probability of victory, justifying the choice for achieving a comprehensive analysis and prediction of the 2024 U.S. Presidential Election.

## Model Summary

### Trump Model Summary
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-fixed-effects
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false
